\bf\mftitle

### 2011-01-22 doc/jInfer/architecture/architecture.tex
removing "schema.data"
3154619 Create documentation for jInfer architecture - better pictures for modules_inference and module_deps (stretching to textwidth with bigger font)
3154619 Create documentation for jInfer architecture - resolved my todos, given to reseto
3159165 Create documentation fof TwoStep simplifier - experimenting
formatting paragraphs and verbatims in architecture

\begin{document}
\maketitle
Target audience: developers willing to extend jInfer.\\

\emph{Note: we use the term \textbf{inference} for the act of creation of schema throughout this and other jInfer documents.}\\

The description of jInfer architecture will be divided into 3 main views: the data view, the process view and the plaform view.\\

The data view will commence by describing the data structures, namely representations of regular expressions and XML elements, attributes and simple data. The notions of a \emph{rule} and a \emph{grammar} will be explained, along with the finite state automata representation we use.\\

Afterwards the process of inference will be described, both from high-level and programmatic point of view.\\

Finally, we will look at jInfer as at an NetBeans Platform application consisting of a number of modules performing their tasks and communicating together through a set of well-defined interfaces.\\

@@

Regular expression is implemented as class \code{Regexp<T>} with supplementing classes \code{RegexpInterval} and \code{RegexpType}.
Each \code{Regexp<T>} instance has one of the enum \code{RegexpType} type:
\begin{itemize}
	\item Lambda ($\lambda$) - empty string (also called $\epsilon$ in literature),
	\item Token - a letter of the alphabet,
	\item Concatenation - one or more regular expression in an ordered sequence. Eg. $(a, b, c, d)$,
	\item Alternation - a choice between one or more regular expressions. Eg. $(a | b | c | d)$,
	\item Permutation - shortcut for all possible permutations of regular expressions. Our syntax to write down permutation is $(a\& b\& c\& d)$.
\end{itemize}

Type of regexp is held in \code{type} member in class \code{Regexp<T>} and can be tested by calling methods \code{isLambda()}, \code{isToken()} etc.

@@

Regexp instance is by default created as immutable, that is, once instantiated, you cannot add more children to list of children, cannot change type, content etc. It is to prevent missuse. In special circumstances, one does not know future children of regexp in time of creation. This occurs mainly in input modules, where by parsing XML data sequentially, one does not know contents of element in time of handling start element event.
For these cases, special \code{getMutable()} method is implemented to obtain regexp with none of members set. One has to fill in all properties carefully and call \code{setImmutable()} aftewards. Proper usage should be one of following:
\begin{verbatim}
	Regexp<T> r = Regexp.<T>getMutable();
	r.setInterval(...);
	r.setType(RegexpType.LAMBDA);
	r.setImmutable();cz.cuni.mff.ksi.jinfer.base
\end{verbatim}
\begin{verbatim}
	Regexp<T> r = Regexp.<T>getMutable();
	r.setInterval(...);
	r.setType(RegexpType.TOKEN);
	r.setContent(...)
	r.setImmutable();
\end{verbatim}
\begin{verbatim}
	Regexp<T> r = Regexp.<T>getMutable();
	r.setInterval(...);
	r.setType(RegexpType.CONCATENATION);
	r.addChild(...);
	r.addChild(...);
	r.addChild(...);
	r.setImmutable();
\end{verbatim}

@@

Finally, our interface/class model for representing XML nodes is drafted on fig. \ref{nodes}. Those, who are brave enough, can look on fig. \ref{nodes_full}.

In result, \code{Element} and \code{SimpleData} have method \code{getType()} to devise type of \code{AbstractStructuralNode} variables. And \code{SimpleData} and \code{Attribute} have methods \code{getContentType()} and \code{getContent()} to work with content model.

@@

\subsection{Rules and grammars} \label{rules_grammar}

jInfer and its documentation uses extended context-free grammars\cite{extendedcfg}.
% TODO vektor check citation of http://www.engr.mun.ca/~theo/Courses/fm/pub/context-free.pdf in literature.bib
% TODO anti what do you mean, "check citation"?
\emph{Rules} in such grammar are in the form
$$
  \textnormal{Left Hand Side (LHS)} \to \textnormal{Right Hand Side (RHS)}
$$
where LHS is a letter of the alphabet (token), RHS is a regular expression over this alphabet. Example would be
$$
  a \to b, (c | d)\ast
$$
In jInfer each such rule is represented with an \code{Element} instance. In this representation, the \code{Element} itself is the LHS, its \code{subnodes} are the RHS.\\

@@

\subsection{Nondeterministic Finite Automaton}
We recommend skipping this section in first read as it describes advanced features.
For all inference algorithms based on merging states of NFA's, our implementation of nondeterministic finite automaton might be interesting.
Implementation consists of 4 classes: \code{Automaton<T>}, \code{Step<T>}, \code{State<T>} and \code{AutomatonCloner<A, B>} (see package \code{cz.cuni.mff.ksi.jinfer.base.automaton}).
Whole implementation uses java generics for representation of symbol of alphabet.
We denote \code{T} the java type of symbol.
Take care of symbol class and its \code{equals()} implementation.
Automaton uses \code{equals()} to compare symbols on transitions (when building prefix-tree automaton and when merging states).
If you are using strings, you're just fine, but with complicated objects, either take care, that equivalent objects are properly tested in \code{equals()}.
Or (maybe faster) solution is to cluster objects into classes of equivalence before inserting them into automaton.
Then give automaton only cluster representant object (which will use java \code{Object.equals()} with reference comparison) for each object encountered.

@@

Automaton supports creating of automaton as a copy of another one (not reference copy, but deep copy expect of symbols), this can be used when searching solution space to create more versions of automaton to edit.
We implemented building of prefix-tree automaton (PTA) in \code{buildPTAOnSymbol()} method.
Create empty automaton and then call this method for every input string of language.
You will get PTA with useCounts and finalCounts set properly on steps/states.

\begin{figure}
	\centering
	\subfigure[Before merge.]
	{
		\includegraphics[scale=\myscale]{automaton_merge1}
		\label{automaton_merge1}
	}
	\subfigure[After states 3 and 4 merged.]
	{
		\centering\includegraphics[scale=\myscale]{automaton_merge2}
		\label{automaton_merge2}
	}
	\subfigure[After states 2 and 4 merged.]
	{
		\centering\includegraphics[scale=\myscale]{automaton_merge3}
		\label{automaton_merge3}
	}
	\caption{Sample automaton.} \label{automaton_merge}
\end{figure}
The biggest thing we offer to scientists is state merging by simply calling \code{mergeStates(state1, state2)} method.
Method merges second state given into first one (or an overloaded version - all states in list into first one in list).
All \{in|out\}-transitions are redirected properly (or discarted as needed).
Variables useCount and finalCount are updated to sums of values from merged transitions/states.
One can ask for merging states, that are merged out from automaton already.
Lets take example automaton of fig. \ref{automaton_merge1} (states are labeled [name|finalCount], steps by \{symbol|useCount\}).
One asks to merge states 3 and 4.
State 3 then becomes final state with loop and state 4 disappears (fig. \ref{automaton_merge2}).
If then one ask to merge states 2 and 4, automaton properly handles situation by knowing, that old state 4, was merged into state 3, and merges states 2 and 3 (see fig. \ref{automaton_merge3}).
This can be useful in $k,h-context$ and $s,k-strings$ implementations (find all context/strings, then supply list of states to merge and don't bother with state names updates).

@@

\section{Process view} \label{section_inference_process}
\begin{figure}
	\centering\includegraphics[scale=\myscale]{inference_process}
	\caption{High-level view of the inference process} \label{inference_process}
\end{figure}

@@

	% TODO rio review this example!

@@

Important thing to note here is that all these steps are executed consecutively. That means, \jmodule{Simplifier} is only started \emph{after} the \jmodule{IGG} completely finished its work and returned IG to be simplified. Similarly, \jmodule{SchemaGen} gets all the rules to export at once, in one list.\\

@@

\subsection{Programmatic view}

From developer's point of view, inference modules are just properly annotated classes (see \ref{section_lookups}) implementing one of the following interfaces (all found under \code{cz.cuni.mff.ksi.jinfer.base.interfaces.inference})

\begin{itemize}
	\item \code{IGGenerator}
	\item \code{Simplifier}
	\item \code{SchemaGenerator}
\end{itemize}

A nice way to name such a class is by adding \code{-Impl} to the name of implemented interface, for example \code{SimplifierImpl}.\\

Annotation required for the framework to recognize such a class as an inference module is the following
\begin{verbatim}
	@ServiceProvider(service = <interface>.class)
\end{verbatim}
for example
\begin{verbatim}
	@ServiceProvider(service = Simplifier.class)
\end{verbatim}

The most important method in each module is \code{start}, defined in each of the interfaces. This method is called by the framework when the respective step of inference is to be executed. It has always two parameters: the actual input data for the module, and a callback object to report to when this step is finished. We will look at both parameters now in more detail.

\subsubsection{Module input}
Each inference module takes the actual input data as the first parameter of its \code{start} method. The type of the argument differs based on the inference module.\\

@@

\jmodule{Initial Grammar Generator} takes an object of type \code{Input}. This class encapsulates all the input files in 3 collections of \code{File}: \code{documents}, \code{schemas} and \code{queries}. Enumerating these files provides IGG with access to all data it needs to create Initial Grammar.\\

\paragraph{Initial Grammar} has two formats:
\begin{itemize}
	\item Only concatenations. This is default format. All regular expressions coming from IGG have to be concatenations, even empty.
	When importing XML documents, this is just fine, as document is positive example from language we are trying to identify.
	But even when importing schemas, queries or other stuff that can contain more information, one has to convert it somehow to concatenations.
	For example, when importing DTD:
	\begin{verbatim}
	<!ELEMENT person (info?, more, more?, more?)>
	<!ELEMENT info (#PCDATA | note)*>
	<!ELEMENT note EMPTY>
	<!ELEMENT more EMPTY>
	\end{verbatim}
	Elements note and more are empty, but we cannot use lambda regexp. Import will use empty concatenation instead.
	For element person, import module have to generate some positive examples of language generated by regular expression \verb+(info?, more, more?, more?)+. This generating is basicly done by calling \code{cz.cuni.mff.ksi.jinfer.base.interfaces.Expander} interface method \code{expand()}. One can implement own intelligent expander, or use our provided \code{cz.cuni.mff.ksi.jinfer.basicigg.expansion.ExpanderImpl}.

	\item Complex regular expressions. Simplifier module running in current inference can indicate support for this format by having capability called \code{can.handle.complex.regexps}.
	Even then, import module is not obliged to use it of course. But it is reasonable to do so, if imported file type gives more information than simple concatenations (when importing schemas).
	Initial Grammar can contain all regular expressions. If schema defines empty element, make its subnodes lambda regexp. If schema defines element content as \verb.(a, b){2,7}(c | d)., just construct that regexp. Empty concatenations are also allowed - as they are positive examples in XML document input. Its simplifier business to divide rules by import file formats (by using metadata described in \ref{section_metada}) and to handle empty concatenations from XML file import with complex regular expressions from other formats.
\end{itemize}
If you are writing import module, check if you can provide more information to simplifier from input file type by using complex format.
This interface division allows us to plug in old methods (accepting only XML input) to the chain of inference, even with schema input files.
It may be useful for method benchmarking too.

@@

\subsubsection{Module output}
Second parameter of each inference module's \code{start} method is a callback object. There are 3 callback interfaces defined in the \code{cz.cuni.mff.ksi.jinfer.base.interfaces.inference} package

\begin{itemize}
	\item \code{IGGeneratorCallback}
	\item \code{SimplifierCallback}
	\item \code{SchemaGeneratorCallback}
\end{itemize}

Each callback interface naturally belongs to the similarly named module interface. As their respective interfaces, also callbacks define one crucial method: \emph{finished}. Each inference module is responsible for invoking this method on the callback it got as a parameter, after it has finished its work and has results to be passed on. Again, these 3 finished methods have different arguments based on the inference module.\\

\code{IGGeneratorCallback.finished()} and \code{SimplifierCallback.finished()} have a grammar (Initial Grammar in case of IGG) as their only argument.\\

@@

\subsubsection{Error handling}

Because the run of each inference module is encapsulated in a \code{try-catch} block by the framework, it is safe to throw any exception out of the \code{start} method: it will get logged, presented to the user and inference will stop. However, if the module uses threads that could throw an exception, it is responsible for catching these exceptions and possibly re-throwing them in the thread where \code{start} runs.

\subsubsection{Interruptions}

User running the inference might change his mind and try to stop this. For this reason, modules have to check for this case in every time-consuming place such as long loops with the following code:
\begin{verbatim}
  for (forever) {
    if (Thread.interrupted()) {
      throw new InterruptedException();
    }
    doStuff();
  }
\end{verbatim}

\subsubsection{Runner} \label{section_runner}

The part of framework responsible for actually gathering user input, running all modules one after another and presenting the results is the \code{Runner} class  in \code{cz.cuni.mff.ksi.jinfer.runner} package.\\

A new instance of \code{Runner} is constructed for each inference run. While being created, \code{Runner}  loads the preferences for current project and looks up user-selected inference modules. Also, callback objects pointing back to methods in \code{Runner} are created. The inference process itself is then as follows

\begin{figure}
\centering\includegraphics[scale=\myscale]{runner}
\caption{Runner} \label{runner}
\end{figure}

\begin{enumerate}
	\item Selected IGG's \code{start} is encapsulated with error/interruption handling and executed, passing \code{Input} and first callback as parameters.
	\item When IGG finishes, it invokes callback's \code{finish} method, passing the IG as parameter.
	\item This in turn causes \code{Runner} to encapsulate and execute Simplifier's \code{start}, passing IG from the first callback and the second callback as parameters.
	\item When Simplifier finishes, it invokes callback's \code{finish} method, passing the simplified grammar as parameter.
	\item This again causes \code{Runner} to encapsulate and execute SchemaGen's \code{start}, passing the simplified grammar from the second callback and the third callback as parameters.
	\item SchemaGen finishes and invokes the last callback's \code{finish}, passing the resulting schema and its extension as parameters.
	\item \code{Runner} receives the resulting schema and based on preferences, saves it to a file, displays it, etc.
\end{enumerate}

\subsection{Node metadata} \label{section_metadata}
To allow simple extensibility of rules, class \code{AbstractNamedNode} contains a string-addressed map called \code{metadata} that can contain arbitrary object values.\\

At the present, jInfer modules use the following metadata, all defined in \code{IGGUtils} class:
\begin{itemize}
	\item \code{from.xml}, filled in by BasicIGG module: means that this rule was created (originates from) from a XML document.
	\item \code{from.schema}, filled in by BasicIGG and XSDImport modules: means that this rule was created (originates from) from a schema.
	\item \code{from.query}, filled in by BasicIGG and XSDImport modules: means that this rule was created from a query.
	\item \code{is.sentinel}, filled in by  BasicIGG and XSDImport module: when importing XML document, one can build whole tree in memory (as in fig. \ref{xml_example}) and then construct list of rules from it (thus saving memory). With schema import however, one not only doesn't know right side of rule in advance (solved by stack in XML import), but right side can be defined anywhere in source file. To save complicated loading of whole schema, searching and pairing elements thorough rules, nodes on right side of rule are created empty - holding only name of node. Fact, that this node has no more information than it's name and position on right side of rule is denominated by labeling the node as sentinel.
	\item \code{schema.data}, filled in by XSDImport module: TODO reseto Explain
\end{itemize}
All of these metadata are of set/not set character (using \code{Boolean.TRUE} as value, but their presence effectively mean true).

\subsection{Capabilities} \label{section_capabilities}
From time to time, modules need a flow of certain information in completely opposite way to the usual inference: they need to know what the following module can or cannot do.\\

For this reason, inference modules extend the \code{Capabilities} interface defining a single method: \code{getCapabilities()}. This method returns a list of strings representing the abilities, or "capabilities" of this module.\\
An inference module can query the capabilities of the \emph{next} module in the inference chain by invoking the 
\begin{verbatim}
RunningProject.getNextModuleCaps()
\end{verbatim}
method to obtain the next module encapsulated in the \code{Capabilities} interface.\\

At the moment, only one capability is defined to be communicated across modules, more specifically between \jmodule{IGGenerator} and \jmodule{Simplifier}. This is \code{Capabilities.CAN\_HANDLE\_COMPLEX\_REGEXPS} meaning whether the \code{Simplifier} accepts more complicated regular expressions than simple concatenations of tokens on its input.\\
% TODO vektor check later - didn't this change?!?

@@

\section{Platform view}

jInfer, being a NetBeans Platform application, is divided into multiple \emph{NetBeans Modules}. It is very important to understand that this division is in principle orthogonal to that naturally sketched in the previous chapters - theoretically, the whole jInfer could be contained in a single NB module, because the framework uses Lookups to locate logical modules. However, for the sake of organization, jInfer is split into several modules and most of the time there is a strong correlation between logical units of jInfer and its NB modules.\\

A short overview of these modules and their functions follows. To get a deeper understanding of any of them, please refer to its documentation. To get deeper understanding of how NBP modules work, their versions, dependencies etc, refer to \cite{modules_api}.\\

Base modules:

\begin{itemize}
	\item \jmodule{jInfer}: not a NBP module in fact, but rather a module suite - the ``root" NB project for the whole framework.
	\item \jmodule{Base}: contains common data structures, interfaces and utility logic shared across all other modules.
\end{itemize}

Inference modules:

\begin{itemize}
	\item \jmodule{Runner}: as described in the section \ref{section_runner}, this module encapsulates the inference process (including the fact that inference consists of 3 phases as described before). Most important class: \code{Runner}.
	\item \jmodule{BasicIGG}: this module contains an extensible implementation of \jmodule{IGGenerator}. Basic version handles all XML-typed input documents, DTD schemas and XPath queries. To find out more about the extensibility to handle other outputs, refer to \jmodule{BasicIGG}'s documentation.
	\item \jmodule{XSDImport}: demonstrating \jmodule{BasicIGG}'s extensibility, this is a base module for XSD input.
	\item \jmodule{XSDImportSAX}: XSD input handling using SAX parser.
	\item \jmodule{XSDImportDOM}: XSD input handling using DOM parser.
	\item \jmodule{TwoStepSimplifier}: a \jmodule{Simplifier} implementation. Implements \cite{ahonen} merging state algorithm with basic handling of attributes. Read the separated documentation of module.
	\item \jmodule{BasicDTDExporter}: this is a \jmodule{SchemaGenerator} implementation with output to DTD.
	\item \jmodule{BasicXSDExporter}: this is a \jmodule{SchemaGenerator} implementation with output to XSD.
\end{itemize}

Utility modules:

>>>>

\begin{document}
\maketitle
\noindent Target audience: developers willing to extend jInfer.

\noindent \emph{Note: we use the term \textbf{inference} for the act of creation of schema throughout this and other jInfer documents.}\\

The description of jInfer architecture will be divided into 3 main views: the data view, the process view and the plaform view.
The data view will commence by describing the data structures, namely representations of regular expressions and XML elements, attributes and simple data. The notions of a \emph{rule} and a \emph{grammar} will be explained, along with the finite state automata representation we use.

Afterwards the process of inference will be described, both from high-level and programmatic point of view.
Finally, we will look at jInfer as at an NetBeans Platform application consisting of a number of modules performing their tasks and communicating together through a set of well-defined interfaces.

@@

Regular expression is implemented as class \code{Regexp<T>} with supplementing classes \code{RegexpInterval} and \code{RegexpType}.
Each \code{Regexp<T>} instance has one of the enum \code{RegexpType} type:
\begin{itemize}
	\item Lambda ($\lambda$) - empty string (also called $\epsilon$ in literature),
	\item Token - a letter of the alphabet,
	\item Concatenation - one or more regular expression in an ordered sequence. Eg. $(a, b, c, d)$,
	\item Alternation - a choice between one or more regular expressions. Eg. $(a | b | c | d)$,
	\item Permutation - shortcut for all possible permutations of regular expressions. Our syntax to write down permutation is $(a\& b\& c\& d)$.
\end{itemize}
Type of regexp is held in \code{type} member in class \code{Regexp<T>} and can be tested by calling methods \code{isLambda()}, \code{isToken()} etc.

@@

For these cases, special \code{getMutable()} method is implemented to obtain regexp with none of members set. One has to fill in all properties carefully and call \code{setImmutable()} aftewards. Proper usage should be one of following:
Regexp instance is by default created as immutable, that is, once instantiated, you cannot add more children to list of children, cannot change type, content etc. It is to prevent missuse. In special circumstances, one does not know future children of regexp in time of creation. This occurs mainly in input modules, where by parsing XML data sequentially, one does not know contents of element in time of handling start element event.
\begin{verbatim}
	Regexp<T> r = Regexp.<T>getMutable();
	r.setInterval(...);
	r.setType(RegexpType.LAMBDA);
	r.setImmutable();cz.cuni.mff.ksi.jinfer.base
\end{verbatim}
\begin{verbatim}
	Regexp<T> r = Regexp.<T>getMutable();
	r.setInterval(...);
	r.setType(RegexpType.TOKEN);
	r.setContent(...)
	r.setImmutable();
\end{verbatim}
\begin{verbatim}
	Regexp<T> r = Regexp.<T>getMutable();
	r.setInterval(...);
	r.setType(RegexpType.CONCATENATION);
	r.addChild(...);
	r.addChild(...);
	r.addChild(...);
	r.setImmutable();
\end{verbatim}

@@

Finally, our interface/class model for representing XML nodes is drafted on fig. \ref{nodes}. Those, who are brave enough, can look on fig. \ref{nodes_full}.
In result, \code{Element} and \code{SimpleData} have method \code{getType()} to devise type of \code{AbstractStructuralNode} variables. And \code{SimpleData} and \code{Attribute} have methods \code{getContentType()} and \code{getContent()} to work with content model.

@@

\subsection{Rules and grammars} \label{rules_grammar}
jInfer and its documentation uses extended context-free grammars\cite{extendedcfg}.
\emph{Rules} in such grammar are in the form
$$
  \textnormal{Left Hand Side (LHS)} \to \textnormal{Right Hand Side (RHS)}
$$
where LHS is a letter of the alphabet (token), RHS is a regular expression over this alphabet. Example would be
$$
  a \to b, (c | d)\ast
$$
In jInfer each such rule is represented with an \code{Element} instance. In this representation, the \code{Element} itself is the LHS, its \code{subnodes} are the RHS.

@@

\subsection{Nondeterministic Finite Automaton}
We recommend skipping this section in first read as it describes advanced features.

For all inference algorithms based on merging states of NFA's, our implementation of nondeterministic finite automaton might be interesting.
Implementation consists of 4 classes (see package \code{cz.cuni.mff.ksi.jinfer.base.automaton}): 
\begin{itemize}
  \item \code{Automaton<T>},
  \item \code{Step<T>},
  \item \code{State<T>},
  \item \code{AutomatonCloner<A, B>}.
\end{itemize}
Whole implementation uses java generics for representation of symbol of alphabet.
We denote \code{T} the java type of symbol.

Take care of symbol class and its \code{equals()} implementation.
Automaton uses \code{equals()} to compare symbols on transitions (when building prefix-tree automaton and when merging states).
If you are using strings, you're just fine, but with complicated objects, either take care, that equivalent objects are properly tested in \code{equals()}.
Or (maybe faster) solution is to cluster objects into classes of equivalence before inserting them into automaton.
Then give automaton only cluster representant object (which will use java \code{Object.equals()} with reference comparison) for each object encountered.

@@

Automaton supports creating of automaton as a copy of another one (not reference copy, but deep copy expect of symbols), this can be used when searching solution space to create more versions of automaton to edit.
We implemented building of prefix-tree automaton (PTA) in \code{buildPTAOnSymbol()} method.
Create empty automaton and then call this method for every input string of language.
You will get PTA with useCounts and finalCounts set properly on steps/states.
\begin{figure}
	\centering
	\subfigure[Before merge.]
	{
		\includegraphics[scale=\myscale]{automaton_merge1}
		\label{automaton_merge1}
	}
	\subfigure[After states 3 and 4 merged.]
	{
		\centering\includegraphics[scale=\myscale]{automaton_merge2}
		\label{automaton_merge2}
	}
	\subfigure[After states 2 and 4 merged.]
	{
		\centering\includegraphics[scale=\myscale]{automaton_merge3}
		\label{automaton_merge3}
	}
	\caption{Sample automaton.} \label{automaton_merge}
\end{figure}
The biggest thing we offer to scientists is state merging by simply calling \code{mergeStates(state1, state2)} method.
Method merges second state given into first one (or an overloaded version - all states in list into first one in list).
All \{in|out\}-transitions are redirected properly (or discarted as needed).
Variables \code{useCount} and \code{finalCount} are updated to sums of values from merged transitions/states.

One can ask for merging states, that are merged out from automaton already.
Lets take example automaton of fig. \ref{automaton_merge1} (states are labeled [name|finalCount], steps by \{symbol|useCount\}).
One asks to merge states 3 and 4.
State 3 then becomes final state with loop and state 4 disappears (fig. \ref{automaton_merge2}).
If then one ask to merge states 2 and 4, automaton properly handles situation by knowing, that old state 4, was merged into state 3, and merges states 2 and 3 (see fig. \ref{automaton_merge3}).
This can be useful in $k,h-context$ and $s,k-strings$ implementations (find all context/strings, then supply list of states to merge and don't bother with state names updates).

@@

\section{Process view} \label{section_inference_process}
\begin{figure}
	\centering\includegraphics[scale=\myscale]{inference_process}
	\caption{High-level view of the inference process} \label{inference_process}
\end{figure}

@@

@@

Important thing to note here is that all these steps are executed consecutively. That means, \jmodule{Simplifier} is only started \emph{after} the \jmodule{IGG} completely finished its work and returned IG to be simplified. Similarly, \jmodule{SchemaGen} gets all the rules to export at once, in one list.

@@

\subsection{Programmatic view}
From developer's point of view, inference modules are just properly annotated classes (see \ref{section_lookups}) implementing one of the following interfaces (all found under \code{cz.cuni.mff.ksi.jinfer.base.interfaces.inference})
\begin{itemize}
	\item \code{IGGenerator}
	\item \code{Simplifier}
	\item \code{SchemaGenerator}
\end{itemize}
A nice way to name such a class is by adding \code{-Impl} to the name of implemented interface, for example \code{SimplifierImpl}.
Annotation required for the framework to recognize such a class as an inference module is the following
\begin{verbatim}
	@ServiceProvider(service = <interface>.class)
\end{verbatim}
for example
\begin{verbatim}
	@ServiceProvider(service = Simplifier.class)
\end{verbatim}
The most important method in each module is \code{start}, defined in each of the interfaces. This method is called by the framework when the respective step of inference is to be executed. It has always two parameters: the actual input data for the module, and a callback object to report to when this step is finished. We will look at both parameters now in more detail.

\subsubsection{Module input}
Each inference module takes the actual input data as the first parameter of its \code{start} method. The type of the argument differs based on the inference module.

@@

\jmodule{Initial Grammar Generator} takes an object of type \code{Input}. This class encapsulates all the input files in 3 collections of \code{File}: \code{documents}, \code{schemas} and \code{queries}. Enumerating these files provides IGG with access to all data it needs to create Initial Grammar.

\paragraph{Initial Grammar} has two formats:
\begin{itemize}
	\item Only concatenations. This is default format. All regular expressions coming from IGG have to be concatenations, even empty.
	When importing XML documents, this is just fine, as document is positive example from language we are trying to identify.
	But even when importing schemas, queries or other stuff that can contain more information, one has to convert it somehow to concatenations.
	For example, when importing DTD:
	\begin{verbatim}
	<!ELEMENT person (info?, more, more?, more?)>
	<!ELEMENT info (#PCDATA | note)*>
	<!ELEMENT note EMPTY>
	<!ELEMENT more EMPTY>
	\end{verbatim}
	Elements note and more are empty, but we cannot use lambda regexp. Import will use empty concatenation instead.
	For element person, import module have to generate some positive examples of language generated by regular expression \verb+(info?, more, more?, more?)+.

	This generating is basicaly done by calling \code{cz.cuni.mff.ksi.jinfer.base.interfaces.Expander} interface method \code{expand()}. One can implement own intelligent expander, or use our provided\\ \code{cz.cuni.mff.ksi.jinfer.basicigg.expansion.ExpanderImpl}.

	\item Complex regular expressions. Simplifier module running in current inference can indicate support for this format by having capability called \code{can.handle.complex.regexps}.
	Even then, import module is not obliged to use it of course. But it is reasonable to do so, if imported file type gives more information than simple concatenations (when importing schemas).

	Initial Grammar can contain all regular expressions. If schema defines empty element, make its subnodes lambda regexp. If schema defines element content as \verb.(a, b){2,7}(c | d)., just construct that regexp. Empty concatenations are also allowed - as they are positive examples in XML document input. Its simplifier business to divide rules by import file formats (by using metadata described in \ref{section_metadata}) and to handle empty concatenations from XML file import with complex regular expressions from other formats.
\end{itemize}
If you are writing import module, check if you can provide more information to simplifier from input file type by using complex format.
This interface division allows us to plug in old methods (accepting only XML input) to the chain of inference, even with schema input files.
It may be useful for method benchmarking too.

@@

\subsubsection{Module output}
Second parameter of each inference module's \code{start} method is a callback object. There are 3 callback interfaces defined in the \code{cz.cuni.mff.ksi.jinfer.base.interfaces.inference} package
\begin{itemize}
	\item \code{IGGeneratorCallback}
	\item \code{SimplifierCallback}
	\item \code{SchemaGeneratorCallback}
\end{itemize}
Each callback interface naturally belongs to the similarly named module interface. As their respective interfaces, also callbacks define one crucial method: \emph{finished}. Each inference module is responsible for invoking this method on the callback it got as a parameter, after it has finished its work and has results to be passed on. Again, these 3 finished methods have different arguments based on the inference module.

\code{IGGeneratorCallback.finished()} and \code{SimplifierCallback.finished()} have a grammar (Initial Grammar in case of IGG) as their only argument.

@@

\subsubsection{Error handling}
Because the run of each inference module is encapsulated in a \code{try-catch} block by the framework, it is safe to throw any exception out of the \code{start} method: it will get logged, presented to the user and inference will stop. However, if the module uses threads that could throw an exception, it is responsible for catching these exceptions and possibly re-throwing them in the thread where \code{start} runs.

\subsubsection{Interruptions}
User running the inference might change his mind and try to stop this. For this reason, modules have to check for this case in every time-consuming place such as long loops with the following code:
\begin{verbatim}
for (forever) {
  if (Thread.interrupted()) {
    throw new InterruptedException();
  }
  doStuff();
}
\end{verbatim}

\subsubsection{Runner} \label{section_runner}
The part of framework responsible for actually gathering user input, running all modules one after another and presenting the results is the \code{Runner} class  in \code{cz.cuni.mff.ksi.jinfer.runner} package.

A new instance of \code{Runner} is constructed for each inference run. While being created, \code{Runner}  loads the preferences for current project and looks up user-selected inference modules. Also, callback objects pointing back to methods in \code{Runner} are created. The inference process itself is then as follows
\begin{figure}
\centering\includegraphics[scale=\myscale]{runner}
\caption{Runner} \label{runner}
\end{figure}
\begin{enumerate}
	\item Selected IGG's \code{start} is encapsulated with error/interruption handling and executed, passing \code{Input} and first callback as parameters.
	\item When IGG finishes, it invokes callback's \code{finish} method, passing the IG as parameter.
	\item This in turn causes \code{Runner} to encapsulate and execute Simplifier's \code{start}, passing IG from the first callback and the second callback as parameters.
	\item When Simplifier finishes, it invokes callback's \code{finish} method, passing the simplified grammar as parameter.
	\item This again causes \code{Runner} to encapsulate and execute SchemaGen's \code{start}, passing the simplified grammar from the second callback and the third callback as parameters.
	\item SchemaGen finishes and invokes the last callback's \code{finish}, passing the resulting schema and its extension as parameters.
	\item \code{Runner} receives the resulting schema and based on preferences, saves it to a file, displays it, etc.
\end{enumerate}

\subsection{Node metadata} \label{section_metadata}
To allow simple extensibility of rules, class \code{AbstractNamedNode} contains a string-addressed map called \code{metadata} that can contain arbitrary object values.

At the present, jInfer modules use the following metadata, all defined in \code{IGGUtils} class:
\begin{itemize}
	\item \code{from.xml}, filled in by BasicIGG module: means that this rule was created (originates from) from a XML document.
	\item \code{from.schema}, filled in by BasicIGG and XSDImport modules: means that this rule was created (originates from) from a schema.
	\item \code{from.query}, filled in by BasicIGG and XSDImport modules: means that this rule was created from a query.
	\item \code{is.sentinel}, filled in by  BasicIGG and XSDImport module: when importing XML document, one can build whole tree in memory (as in fig. \ref{xml_example}) and then construct list of rules from it (thus saving memory). With schema import however, one not only doesn't know right side of rule in advance (solved by stack in XML import), but right side can be defined anywhere in source file. To save complicated loading of whole schema, searching and pairing elements thorough rules, nodes on right side of rule are created empty - holding only name of node. Fact, that this node has no more information than it's name and position on right side of rule is denominated by labeling the node as sentinel.
\end{itemize}
All of these metadata are of set/not set character (using \code{Boolean.TRUE} as value, but their presence effectively mean true).

\subsection{Capabilities} \label{section_capabilities}
From time to time, modules need a flow of certain information in completely opposite way to the usual inference: they need to know what the following module can or cannot do.

For this reason, inference modules extend the \code{Capabilities} interface defining a single method: \code{getCapabilities()}. This method returns a list of strings representing the abilities, or "capabilities" of this module.
An inference module can query the capabilities of the \emph{next} module in the inference chain by invoking the 
\begin{verbatim}
RunningProject.getNextModuleCaps()
\end{verbatim}
method to obtain the next module encapsulated in the \code{Capabilities} interface.

At the moment, only one capability is defined to be communicated across modules, more specifically between \jmodule{IGGenerator} and \jmodule{Simplifier}. This is \code{Capabilities.CAN\_HANDLE\_COMPLEX\_REGEXPS} meaning whether the \code{Simplifier} accepts more complicated regular expressions than simple concatenations of tokens on its input.
% TODO vektor check later - didn't this change?!?

@@

\section{Platform view}
jInfer, being a NetBeans Platform application, is divided into multiple \emph{NetBeans Modules}. It is very important to understand that this division is in principle orthogonal to that naturally sketched in the previous chapters - theoretically, the whole jInfer could be contained in a single NB module, because the framework uses Lookups to locate logical modules. However, for the sake of organization, jInfer is split into several modules and most of the time there is a strong correlation between logical units of jInfer and its NB modules.

A short overview of these modules and their functions follows. To get a deeper understanding of any of them, please refer to its documentation. To get deeper understanding of how NBP modules work, their versions, dependencies etc, refer to \cite{modules_api}.

\noindent Base modules:
\begin{itemize}
	\item \jmodule{jInfer}: not a NBP module in fact, but rather a module suite - the ``root" NB project for the whole framework.
	\item \jmodule{Base}: contains common data structures, interfaces and utility logic shared across all other modules.
\end{itemize}

\noindent Inference modules:
\begin{itemize}
	\item \jmodule{Runner}: as described in the section \ref{section_runner}, this module encapsulates the inference process (including the fact that inference consists of 3 phases as described before). Most important class: \code{Runner}.
	\item \jmodule{BasicIGG}: this module contains an extensible implementation of \jmodule{IGGenerator}. Basic version handles all XML-typed input documents, DTD schemas and XPath queries. To find out more about the extensibility to handle other outputs, refer to \jmodule{BasicIGG}'s documentation.
	\item \jmodule{XSDImport}: demonstrating \jmodule{BasicIGG}'s extensibility, this is a base module for XSD input.
	\item \jmodule{XSDImportSAX}: XSD input handling using SAX parser.
	\item \jmodule{XSDImportDOM}: XSD input handling using DOM parser.
	\item \jmodule{TwoStepSimplifier}: a \jmodule{Simplifier} implementation. Implements \cite{ahonen} merging state algorithm with basic handling of attributes. Read the separated documentation of module.
	\item \jmodule{BasicDTDExporter}: this is a \jmodule{SchemaGenerator} implementation with output to DTD.
	\item \jmodule{BasicXSDExporter}: this is a \jmodule{SchemaGenerator} implementation with output to XSD.
\end{itemize}
